Environmental Context Dependency in Species Interactions

### 2018-08-23 EDM_SNI_analysis.Rmd
manuscript and figure edits

# label with the number of incoming(outgoing) connections
vert_in <- sapply(igraph::degree(g2,mode="in"),function(x) paste0("(",x,")"))
vert_out_in <- paste0(igraph::degree(g2,mode="out"),vert_in)
V(g2)$label <- paste(sep="\n",V(g2)$name, vert_out_in)
V(g2)$label[7:11] <- paste(V(g2)$name[7:11])

@@

#plot to file if uncommented
# png("plots/ccm_network.png",height=500,width=600)
plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
# dev.off()
```

@@

```{r mean vs variance}
int_mean_sd <-spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  group_by(mod,effect,type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),
            cv_int=sd_int/abs_mean*100,upper=quantile(value,0.975),lower=quantile(value,0.025),
            perc_neutral=sum(neutral)/n()*100)%>%
  ungroup()

# Mean interaction strength by interaction type
spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  filter(mod != "ymac",effect != 'ymac')%>%
  group_by(type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),n=n(),upper=quantile(value,0.975),lower=quantile(value,0.025))%>%
  ungroup()%>%
  ggplot(aes(type,mean_int))+
  geom_col()+
  geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
  labs(x='Interaction Type',y='Interaction Strength')

>>>>

# label with the number of incoming(outgoing) connections
# vert_in <- sapply(igraph::degree(g2,mode="in"),function(x) paste0("(",x,")"))
# vert_out_in <- paste0(igraph::degree(g2,mode="out"),vert_in)
# V(g2)$label <- paste(sep="\n",V(g2)$name, vert_out_in)
# V(g2)$label[7:11] <- paste(V(g2)$name[7:11])

@@

#plot to file if uncommented
# png("plots/ccm_network.png",height=500,width=600)
# pdf("plots/ccm_network.pdf")
plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.8, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
# dev.off()
```

@@

```{r mean vs variance}
int_mean_sd <-spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  group_by(mod,effect,type) %>%
  summarise(median_int=median(value,na.rm=T),mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),
            cv_int=sd_int/abs_mean*100,upper=quantile(value,0.975),lower=quantile(value,0.025),
            perc_neutral=sum(neutral)/n()*100)%>%
  ungroup()

# Mean interaction strength by interaction type
spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  filter(mod != "ymac",effect != 'ymac')%>%
  group_by(type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),n=n(),upper=quantile(value,0.975),lower=quantile(value,0.025))%>%
  ungroup()%>%
  ggplot(aes(type,mean_int))+
  geom_col()+
  # geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
  labs(x='Interaction Type',y='Interaction Strength')

# Median interaction strength by interaction type
spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  filter(mod != "ymac",effect != 'ymac')%>%
  group_by(type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),median_int=median(value,na.rm=T),n=n(),upper=quantile(value,0.975),lower=quantile(value,0.025))%>%
  ungroup()%>%
  ggplot(aes(type,median_int))+
  geom_col(fill="gray60")+
  geom_point(data=int_mean_sd,aes(type,median_int,color=type),position="dodge")+
  scale_color_npg()+
  guides(color=F)+
  # geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
  labs(x='Interaction Type',y='Median Interaction Strength')

### 2018-08-23 manuscript.Rmd
manuscript and figure edits

---
title: "Environmental Context Dependency in Species Interactions"
author: "Owen Liu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
fontsize: 12pt
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \renewcommand\linenumberfont{\normalfont\bfseries\small}
csl: nature.csl
bibliography: citation_bibtex.bib
---

@@

Recently-developed EDM methods exist for uncovering dynamic species interactions from time series data [@Deyle2016], but these methods have insofar been applied only to simulated and planktonic communities, and their utility to the study of other ecological systems remains untested. We focus here instead on giant kelp forests in southern California, a diverse and temporally dynamic ecosystem in which many important species interactions are well-documented [@Foster1985a; @Graham2007c; @Teagle2017]. The study of kelp forests has been foundational to ecological theory, especially regarding the relative influence of top-down and bottom-up structuring forces in ecosystems [@Estes1974a; @Graham1997; @Kvitek1998a; @Foster2010; @Reed2011]. Recently, however, findings from long-term kelp forest research programs have begun to challenge many long-held beliefs about the drivers of kelp forest ecosystem dynamics [@Reed2016]. In particular, a longer-term perspective has led to a recognition of the critical importance of environmental context—such as level of physical disturbance or the current state of El Niño conditions—for understanding kelp forest processes [@Parnell2010; @Byrnes2011; @Bell2015; @Young2015]. In this study we utilize monitoring data from one such effort at San Nicolas Island, a small, remote member of the California Channel Islands in the northeast Pacific [@Kenner2013].

@@

Applying convergent cross mapping[@Sugihara2012] (CCM) to the set of six biological and five physical variables, we find a relatively dense interaciton network (Figure \ref{fig:ccm}). Out of 90 possible unidirectional links between species and between species and the environmental variables, 40 are significant. Adult *Macrocystis* density is driven by all five environmental variables, with SWH, SST, and the NPGO showing the strongest causal signals. This finding confirms recent work by others using different methods [@Bell2015] that showed that these same three variables were the primary controls of giant kelp biomass dynamics across the California coast. More generally, although the included physical variables show significant links to many of the biological variables, the NPGO and SWH show the strongest links to almost all of the biological variables. Our analysis suggests that more attention should be focused on the effects of the NPGO.

```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping. Each arrow represents a significant causal signal, and link width and opacity scale with the strength of causal forcing (see full results in Supplementary material). Species abbreviations: M. pyr: \\textit{Macrocystis pyrifera}; L.far: \\textit{Laminaria farlowii}; P.cal: \\textit{Pterygophora californica}; M.fra: \\textit{Mesocentrotus franciscanus}; S.pur: \\textit{Strongylocentrotus purpuratus}. Physical drivers: NPGO: North Pacific Gyre Oscillation; MEI: Multivariate El Nino Index; PDO: Pacific Decadal Oscillation; SST: Sea surface temperature; SWH: Significant wave height.\\label{fig:ccm}"}
include_graphics("plots/ccm_network.pdf")
```

Interestingly, in turn, adult *Macrocystis* shows strong causal links to every other biological variable. This is despite the fact that the study site at San Nicolas Island does not have a stable giant kelp forest (see raw time series, Figure \ref{fig:normts}); rather, the site has transitioned from an urchin barren[@Harrold1985a] to a *Pterygophora* and *Laminaria*-dominated state, to a *Macrocystis* forest at various times throughout the 30-year time series. The implication is that, despite not maintaining dominance in the typical ecological sense of word (large abundance and biomass), *Macrocystis* is "dynamically dominant", and remains the key foundation species in this ecosystem. Its dynamics are fundamentally important in driving the dynamics of all the other kelp forest species[@Graham2007c].

CCM analysis confirms that the system studied represents a complex array of significant interactions between algal species and their herbivores. However, CCM alone does not elucidate the direction and magnitude of species interactions. To obtain estimates of the interactions themselves, we use multivariate S-maps[@Deyle2016; @Sugihara1994] (see Methods). S-maps reconstructs dynamic “attractors” by casting the abundances of causally-related species into state space. For a set of causally-related species, a point in multivariate space can be plotted using each species’ abundance as an axis. The attractor is then created by tracing this multispecies trajectory forward in time (see example attractor in Figure \ref{fig:attractor}). For each point along the attractor, S-maps computes a Jacobian matrix, the elements of which are the estimated partial derivatives between species. These interaction matrix elements are our measure of species interactions. Because Jacobians are computed sequentially for every point along reconstructed attractors, we obtain estimates of interaction strength that vary with ecosystem state.

There are two important aspects of S-maps that deserve mention. First, the S-map estimation procedure, like all EDM methods, is specifically designed for nonlinear systems, and is therefore an appropriate tool for investigating ecosystems exhibiting nonlinear dynamics such as alternative stable states or hysteresis[@Deyle2016]. Secondly, because S-maps utilizes reconstructed multispecies attractors, each estimated interaction is fundamentally based on observations of similar past ecosystem states---in our case represented as multivariate collections of causally-linked species' densites---rather than a phenomenological extrapolation of the most recent dynamics. In simple terms, instead of asking, "What is our prediction for the strength of herbivory based on last year's observed dynamics," S-maps is concerned with, "What is our best estimate for the strength of herbivory, based on our knowledge of times in the past when the ecosystem was most similar to today?"

```{r fig.cap="Smoothed kernel density histograms of estimated interactions by type. Solid lines denote means across all estimated interactions of that type.\\label{fig:int_dist}"}
interaction_distribution_by_type
```

In the San Nicolas kelp forest, we find a striking prevalence of neutral and positive species interactions (Figure \ref{fig:int_dist}). After grouping species interactions by type, only herbivory (the effect of urchins on algae species) is predominantly negative. Conversely, the effect of the algae species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions. Likewise, contrary to our expectations, interactions between the algae species and between the urchin species are not always antagonistic. These results suggest that facilitation---direct or indirect---can arise in kelp forests in multiple contexts[@Bennett2015]. Since there is evidence here of strong herbivory, for example, then the indirect facilitative effects between algae species through shared herbivory avoidance[@Harris1984] may sometimes outweigh their direct competition. Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c]. These types of indirect facilitation in kelp forests have received relatively little attention[@Arkema2009a; @Benes2015] relative to a strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other systems[@Baraza2006]. Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an additional important structuring force.

```{r fig.cap="Mean and standard deviation of each estimated unidirectional species interaction (excluding intraspecies effects)\\label{fig:int_mean_range}"}
int_mean_range_plot
```

There is an inverse relationship between mean and range in interaction strengths (Figure \ref{fig:int_mean_range}). Importantly, these reported ranges in interaction strength are distinct from estimation or experimental errors. They represent observed ranges across all individually-estimated S-map interactions from each reconstructed attractor.  While inverse mean-range relationship was true overall, it varied by interaction type. For the three significant interactions between algae species (excluding interactions with *Macrocytis* recruits), the effects with the smallest mean strength had the largest range across the time series. The effects of algal on urchin species were the most variable in general, and the herbivory effects the least. Additionally, the range was greater than the mean for all of these interactions, further supporting the idea that context dependency may be the rule rather than the exception in this ecosystem. The observed mean-range relationship echoes work done across a range of experimental systems[@Berlow1999b], and implies that interactions that seem weak on average may simply vary in sign and magnitude across contexts but still be important in each of those contexts nonetheless.

```{r fig.cap="Distributions of \\textit{Macrocystis} effects on \\textit{Pterygophora}, under high (greater than 1) and low (less than -1) values of five normalized environmental indices (abbreviations as in Figure \\ref{fig:ccm}. Solid lines: mean interactions under each regime.\\label{fig:mac_pter}"}
mac_effect_on_pter_plot
```

>>>>

---
title: "Environmental Context Dependency in Species Interactions"
author: "Owen Liu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: inline
fontsize: 12pt
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \renewcommand\linenumberfont{\normalfont\bfseries\small}
csl: nature.csl
bibliography: citation_bibtex.bib
---

@@

Recently-developed EDM methods exist for uncovering dynamic species interactions from time series data [@Deyle2016], but these methods have to date been applied only to simulated and planktonic communities, and their utility to the study of other ecological systems remains untested. Here we extend the exploration of dynamic species interactions using EDM to giant kelp forests in southern California, a diverse and temporally dynamic ecosystem in which many important species interactions are well-documented through experimental and comparative studies [@Foster1985a; @Graham2007c; @Teagle2017]. The study of kelp forests has been foundational to ecological theory, especially regarding the relative influence of top-down and bottom-up structuring forces in ecosystems [@Estes1974a; @Graham1997; @Kvitek1998a; @Foster2010; @Reed2011]. Recently, however, findings from long-term kelp forest research programs have begun to challenge many long-held beliefs about the drivers of kelp forest ecosystem dynamics [@Reed2016]. In particular, a longer-term perspective has led to a recognition of the critical importance of environmental context—such as level of physical disturbance or the current state of El Niño conditions—for understanding kelp forest processes [@Parnell2010; @Byrnes2011; @Bell2015; @Young2015]. In this study we utilize monitoring data from one such effort at San Nicolas Island, a small, remote member of the California Channel Islands in the northeast Pacific [@Kenner2013].

@@

Applying convergent cross mapping[@Sugihara2012] (CCM) to the set of six biological and five physical variables, we find a relatively dense interaciton network (Figure \ref{fig:ccm}). Out of 90 possible unidirectional links between species and between species and the environmental variables, 40 are significant. Adult *Macrocystis* density is driven by all five environmental variables, with SWH, SST, and the NPGO showing the strongest causal signals. This finding aligns with recent work by others using different methods [@Bell2015] that showed that these same three variables were the primary controls of giant kelp biomass dynamics across the California coast. More generally, although the included physical variables show significant links to many of the biological variables, the NPGO and SWH show the strongest links to almost all of the biological variables. Our analysis suggests that more attention should be focused on the effects of the NPGO.

```{r fig.width=4, fig.height=4,fig.cap="Reconstructed interaction web using results of convergent cross mapping. Each arrow represents a significant causal signal, and link width and opacity scale with the strength of causal forcing (see full results in Supplementary material). Species abbreviations: M. pyr: \\textit{Macrocystis pyrifera}; L.far: \\textit{Laminaria farlowii}; P.cal: \\textit{Pterygophora californica}; M.fra: \\textit{Mesocentrotus franciscanus}; S.pur: \\textit{Strongylocentrotus purpuratus}. Physical drivers: NPGO: North Pacific Gyre Oscillation; MEI: Multivariate El Nino Index; PDO: Pacific Decadal Oscillation; SST: Sea surface temperature; SWH: Significant wave height.\\label{fig:ccm}"}
include_graphics("plots/ccm_network_w_icons.pdf")
```

Interestingly, in turn, adult *Macrocystis* shows strong causal links to every other biological variable. This is despite the fact that the study site at San Nicolas Island does not have a stable giant kelp forest (see raw time series, Figure \ref{fig:normts}); rather, the site has transitioned from an urchin barren[@Harrold1985a] to a *Pterygophora* and *Laminaria*-dominated state, to a *Macrocystis* forest at various times throughout the 30-year time series. The implication is that, despite not maintaining dominance in the typical ecological sense of word (large abundance and biomass), *Macrocystis* remains a key foundation species in this ecosystem, because its dynamics are fundamentally important in driving the dynamics of all the other kelp forest species[@Graham2007c].

CCM analysis confirms that the system studied represents a complex array of significant interactions between algal species and their herbivores. However, CCM alone does not elucidate the direction and magnitude of species interactions. To obtain estimates of the interactions themselves, we use multivariate S-maps[@Deyle2016; @Sugihara1994] (see Methods). S-maps reconstruct dynamic “attractors” by casting the abundances of causally-related species into state space. For a set of causally-related species, a point in multivariate space can be plotted using each species’ abundance as an axis. The attractor is then created by tracing this multispecies trajectory forward in time (see example attractor in Figure \ref{fig:attractor}). For each point along the attractor, S-maps computes a Jacobian matrix, the elements of which are the estimated partial derivatives between species. These interaction matrix elements are our measure of species interactions. Because Jacobians are computed sequentially for every point along reconstructed attractors, we obtain estimates of interaction strength that vary with ecosystem state.

Dynamic ecosystems are analogous to a landscape of variable topography, where our position on the landscape represents the current ecosystem state. In this analogy, each cardinal direction represents the density of a different species (i.e., axis in state-space), and our movement across the landscape through time represents movement along the multivariate attractor as species densities change.  S-maps are simply our reconstructed topographic maps of these ecosystem landscapes. At a given point in time as we move across the landscape, we may be on top of a steep pinnacle, on a flat plain, or in a shallow valley. Regardless, it is the slope of the landscape in each direction (the partial derivative) that defines the local interaction strength between species: a steep upward slope represents a strong positive interaction between species, while a flat surface is a neutral interaction and a slight downward slope is a weak negative interaction.

There are two important characteristics of S-maps that deserve mention. First, the S-map estimation procedure, like all EDM methods, is specifically designed for nonlinear systems, and is therefore an appropriate tool for investigating ecosystems exhibiting nonlinear dynamics such as alternative stable states or hysteresis[@Deyle2016]. Secondly, because S-maps utilizes reconstructed multispecies attractors, each estimated interaction is fundamentally based on observations of similar past ecosystem states---where each state is represented as a multivariate vector of causally-linked species' densites---rather than a phenomenological extrapolation of the most recent dynamics. For example, instead of asking, "What is our prediction for the strength of herbivory based on last year's observed dynamics," S-maps is concerned with, "What is our best estimate for the strength of herbivory, based on our knowledge of times in the past when the ecosystem was most similar to today?"

```{r fig.cap="Smoothed kernel density histograms of estimated interactions by type. Solid lines denote means across all estimated interactions of that type.\\label{fig:int_dist}"}
include_graphics("plots/int_dist_by_type_density.pdf")
# interaction_distribution_by_type
```

In the San Nicolas kelp forest, we find a striking prevalence of neutral and positive species interactions (Figure \ref{fig:int_dist}). After grouping species interactions by type, only herbivory (the effect of urchins on algal species) is predominantly negative. Conversely, the effect of the algal species on the urchins has the flattest distribution, with occasional strong negative and strong positive interactions. Likewise, contrary to our expectations, interactions between the algal species and between the urchin species are not always antagonistic. These results suggest that facilitation---direct or indirect---can arise in kelp forests in multiple contexts[@Bennett2015]. For example, since there is evidence here of strong herbivory, then there may be indirect facilitation between algal species because of a shared herbivore[@Harris1984]: a greater algal density in general could ameliorate the negative effect of herbivory on any one species. This sort of indirect, associative facilitation may sometimes outweigh the strength of direct algal competition. Additionally, *Macrocystis* forests can mediate current strength and as a result, can also help retain reproductive propagules of other species[@Graham2007c]. These types of indirect facilitation in kelp forests have received comparatively little attention[@Arkema2009a; @Benes2015] relative to the strong focus on exploitative competition between these species for light and nutrients[@Reed1984; @Dayton1984], but similar effects have been documented in other ecosystems[@Baraza2006]. Our analysis does not contradict the importance of competition in kelp forest ecosystems, but suggests that facilitation, especially indirect facilitation, may be an additional important structuring force.

```{r fig.cap="Mean absolute interaction strength and standard deviation for each estimated unidirectional species interaction, excluding intraspecies effects\\label{fig:int_mean_range}"}
int_mean_range_plot
```

The empirical distributions in Figure \ref{fig:int_dist} are grouped by interaction type. We can also investigate the distribution of each individual species interaction (e.g., the effect of purple urchins on giant kelp). Between the five unique species, there were 14 significant unidirectional causal links (Figure \ref{fig:ccm}). Because S-maps estimate interactions for each historical ecosystem state, those 14 interactions each have their own distribution of interactions strengths (Figure \ref{fig:int_mean_range}). In general, interactions with larger mean strength (in absolute terms) also had a smaller range. Importantly, these reported ranges in interaction strength are distinct from estimation or experimental errors. They represent observed ranges across all individually-estimated S-map interactions from each reconstructed attractor.  While inverse mean-range relationship was true overall, it varied by interaction type. For the three significant interactions between algal species (excluding interactions with *Macrocytis* recruits), the effects with the smallest mean strength had the largest range across the time series. The effects of algal on urchin species were the most variable in general, and the herbivory effects the least. Additionally, the range was greater than the mean for all of these interactions, further supporting the idea that context dependency may be the rule rather than the exception in this ecosystem. The observed mean-range relationship echoes work done across a range of experimental systems[@Berlow1999b], and implies that interactions that seem weak on average may simply vary in sign and magnitude across contexts but still be important in each of those contexts nonetheless.

```{r fig.cap="Distributions of \\textit{Macrocystis} effects on \\textit{Pterygophora}, under high (greater than 1) and low (less than -1) values of five normalized environmental indices (abbreviations as in Figure \\ref{fig:ccm}. Solid lines: mean interactions under each regime.\\label{fig:mac_pter}"}
# mac_effect_on_pter_plot
include_graphics("plots/mac_effect_on_pter_phys_w_icons.pdf")
```
